Statistical analysis of periodic data in neuroscience

### 2020-12-26 manuscript/manuscript.Rmd
Adding skeleton documentation and package structure

# reasonably compact code to check which packages are installed, 
# install the missing ones, and activate all
packagelist <- c('bookdown','pals','shape','diagram','HDMD')
missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
if (length(missingpackages)>0){install.packages(missingpackages)}
toinstall <- packagelist[which(!packagelist %in% (.packages()))]
invisible(lapply(toinstall,library,character.only=TRUE))

@@

tsq1.test <- function(data,mu=NULL){
  # function to calculate a one-sample t-squared test after Hotelling (1931)
  # for two-sample tests, use the Hotelling package

  # compare each DV to some value
  if (!is.null(mu)){for (m in 1:length(mu)){data[,m] <- data[,m]-mu[m]}}

  xbar <- colMeans(data) # estimate means for each DV
  m <- ncol(data)   # record number of DVs
  N <- nrow(data)   # record sample size
  C <- cov(data)    # calculate covariance matrix
  Cinv <- solve(C)  # calculate inverse matrix
  tsq <- N*t(xbar)%*%Cinv%*%xbar # compute t2 statistic
  Fratio <- tsq * (N-m)/(m*(N-1)) # convert to F-ratio
  df1 <- m      # record degrees of freedom
  df2 <- N-m
  pval <- 1- pf(Fratio,df1,df2) # estimate p-value
  p.value <- min(pval,1)

  # store outputs in a data structure
  output <- data.frame(tsq,Fratio,df1,df2,p.value)
  return(output)}

tsqc.test <- function(x,y=NULL,paired=FALSE,mu=NULL){
  # function to calculate the t-squared circ statistic from Victor & Mast (1991)
  # the inputs must be Nx2 or 2xN matrices of numbers
  # for a one-sample test, mu is an optional vector to which the data are compared

  # transpose if dimensions are the wrong way round
  if (is.matrix(x)){d <- dim(x)
  if (d[1]<d[2]){x <- t(x)}}   
  if (!is.null(y)){if (is.matrix(y)){d <- dim(y)
  if (d[1]<d[2]){y <- t(y)}}}

  if (!is.null(mu)){for (n in 1:length(mu)){x[,n] <- x[,n] - mu[n]}}

  # convert data to complex representation
  x <- complex(real=x[,1],imaginary=x[,2])
  if (!is.null(y)){y <- complex(real=y[,1],imaginary=y[,2])}

  if (is.null(y)){method <- 'One-sample t-squared circ test'}
  if (!is.null(y) & paired==TRUE){method <- 'Paired t-squared circ test'}
  if (!is.null(y) & paired==FALSE){method <- 'Independent samples t-squared circ test'}

  if (paired==TRUE){if (is.null(y)==FALSE){x <- x - y}}

  # paired or one-sample version of the test
  if (paired==TRUE | is.null(y)){

    nobs <- length(x)
    cohmean <- mean(x)
    absmean <- abs(cohmean)

    displacement <- abs(x - cohmean)
    diffsumsq <- sum(displacement^2)
    tsqc <- (nobs-1) * (absmean^2) / diffsumsq
    Fratio <- nobs*tsqc
    df2 <- nobs*2-2
    pval <- pf(Fratio,df1=2,df2=df2,lower.tail=FALSE)
  }

  # independent samples version of the test
  if (paired==FALSE & !is.null(y)){  
    nobs1 <- length(x)
    cohmean1 <- mean(x)
    absmean1 <- abs(cohmean1)
    displacement1 <- abs(x - cohmean1)

    nobs2 <- length(y)
    cohmean2 <- mean(y)
    absmean2 <- abs(cohmean2)    
    displacement2 <- abs(y - cohmean2)

    meandiff <- cohmean1 - cohmean2
    absdiff <- abs(meandiff)
    diffsumsq <- sum(displacement1^2) + sum(displacement2^2)

    tsqc <- (nobs1 + nobs2 - 2) * (absdiff^2) / diffsumsq
    Fratio <- ((nobs1*nobs2)/(nobs1 + nobs2))*tsqc
    df2 <- (2*nobs1 + 2*nobs2 - 4)
    pval <- pf(Fratio,df1=2,df2=df2,lower.tail=FALSE)
  }

  df1 <- 2
  p.value <- min(pval,1)

  output <- data.frame(tsqc,Fratio,df1,df2,p.value,method)
  return(output)}

CI.test <- function(data){
  # function to calculate the condition index of bivariate data
  # compare for significance with an expected distribution function

  C <- cov(data)  # covariance matrix
  N <- nrow(data) # sample size
  eigVal  <- eigen(C)$values  # calculate eigenvalues
  CI <- sqrt(eigVal[1]/eigVal[2]) # calculate condition index

  maxci <- max(20,round(CI+5))   # make sure our target CI is in range
  cilist <- seq(1,maxci,0.01)    # list of possible CIs

  # calculate probability density function using modified Edelman equation
  pdffunction <- ((N-2)*(2^(N-2))) * 
    ((cilist^2 - 1)/((cilist^2+1)^(N-1))) * (cilist^(N-3))
  cdfinverse <- 1-(cumsum(pdffunction)/sum(pdffunction)) # inverse of cdf
  criticalCI <- cilist[min(which(cdfinverse<=0.05))]  # find the threshold CI

  indices <- which(cilist>=CI)  # find the CI values larger than our CI
  pval <- cdfinverse[indices[1]] # estimate the p-value

  output <- data.frame(CI,N,criticalCI,pval)
  return(output)}

anovacirc.test <- function(data){
  # analysis of variance using complex data
  # an extension of the logic of the t-squared circ statistic

  grouplabels <- data$grouplabels
  datavals <- data$simdata
  factorlist <- levels(grouplabels)

  grandmean <- mean(datavals)

  groupmeans <- NULL
  for (n in 1:nlevels(grouplabels)){groupmeans[n] <- mean(datavals[which(grouplabels==factorlist[n])])}

  SSM <- 0
  for (n in 1:nlevels(grouplabels)){SSM <- SSM + length(which(grouplabels==factorlist[n]))*abs(groupmeans[n]-grandmean)^2}

  dfM <- 2*(nlevels(grouplabels)-1)

  allresiduals <- 1:length(datavals)
  for (n in 1:nlevels(grouplabels)){allresiduals[which(grouplabels==factorlist[n])] <- abs(datavals[which(grouplabels==factorlist[n])] - groupmeans[n])}

  SSR <- sum(allresiduals^2)
  dfR <- 2*(length(datavals)-nlevels(grouplabels))
  MSM <- SSM/dfM
  MSR <- SSR/dfR
  Fratio <- MSM/MSR
  pvalue <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)

  output <- NULL
  output$Fratio <- Fratio
  output$pvalue <- pvalue
  output$SSM <- SSM
  output$SSR <- SSR
  output$dfM <- dfM
  output$dfR <- dfR
  output$MSM <- MSM
  output$MSR <- MSR

  return(output)
}

anovacircR.test <- function(data){
  # analysis of variance using complex data
  # an extension of the logic of the t-squared circ statistic
  # this version for repeated measures designs

  grouplabels <- data$grouplabels
  factorlist <- levels(grouplabels)
  participantlabels <- data$participant
  participantlist <- levels(participantlabels)
  datavals <- data$simdata

  grandmean <- mean(datavals)

  SSW <- 0
  for (n in 1:nlevels(participantlabels)){SSW <- SSW + sum(abs(datavals[which(participantlabels==participantlist[n])] - mean(datavals[which(participantlabels==participantlist[n])]))^2)}
  dfW <- 2*(nlevels(participantlabels)*(nlevels(grouplabels)-1))

  groupmeans <- NULL
  for (n in 1:nlevels(grouplabels)){groupmeans[n] <- mean(datavals[which(grouplabels==factorlist[n])])}
  SSM <- 0
  for (n in 1:nlevels(grouplabels)){SSM <- SSM + length(which(grouplabels==factorlist[n]))*abs(groupmeans[n]-grandmean)^2}
  dfM <- 2*(nlevels(grouplabels)-1)

SSR <- SSW - SSM
dfR <- dfW - dfM

  MSM <- SSM/dfM
  MSR <- SSR/dfR
  Fratio <- MSM/MSR
  pvalue <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)

  output <- NULL
  output$Fratio <- Fratio
  output$pvalue <- pvalue
  output$SSW <- SSW  
  output$SSM <- SSM
  output$SSR <- SSR
  output$dfW <- dfW
  output$dfM <- dfM
  output$dfR <- dfR
  output$MSM <- MSM
  output$MSR <- MSR

  return(output)
}

getel <- function(compdata){
  A <- cov(compdata)
  ctr    <- colMeans(compdata) 
  RR     <- chol(A)                               # Cholesky decomposition
  angles <- seq(0, 2*pi, length.out=200)          # angles for ellipse
  ell    <- 1 * cbind(cos(angles), sin(angles)) %*% RR  # ellipse scaled with factor 1
  ellCtr <- sweep(ell, 2, ctr, "+")               # center ellipse to the data centroid
  eigVal  <- eigen(A)$values
  eigVec  <- eigen(A)$vectors
  eigScl  <- eigVec  %*% diag(sqrt(eigVal))  # scale eigenvectors to length = square-root
  xMat    <- rbind(ctr[1] + eigScl[1, ], ctr[1] - eigScl[1, ])
  yMat    <- rbind(ctr[2] + eigScl[2, ], ctr[2] - eigScl[2, ])
  ellBase <- cbind(sqrt(eigVal[1])*cos(angles), sqrt(eigVal[2])*sin(angles)) 
  ellRot  <- eigVec %*% t(ellBase) 
  return(ellRot)
  }

>>>>

# reasonably compact code to check which packages are installed, 
# install the missing ones, and activate all

# devtools::install_github("bakerdh/FourierStats")
# devtools::install('FourierStats')

packagelist <- c('bookdown','pals','shape','diagram','HDMD','FourierStats')
missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
if (length(missingpackages)>0){install.packages(missingpackages)}
toinstall <- packagelist[which(!packagelist %in% (.packages()))]
invisible(lapply(toinstall,library,character.only=TRUE))

@@
